Towards Practical First-Order Model Counting

### 2025-08-08 doc/talk/talk.tex
Ditto

\begin{frame}{Motivation}
  \begin{exampleblock}{Example Setting}
    \begin{itemize}
      \item Let \structure{$\Delta$} be a set of cardinality
            \structure{$n \in \mathbb{N}_0$}
      \item Suppose we want to count all \structure{$P \subseteq \Delta^{2}$}
            (as a function of \structure{$n$}) that are:
            \begin{itemize}
              \item functions,
              \item bijections,
              \item partial orders,
              \item symmetric,
              \item transitive,
              \item etc.
            \end{itemize}
    \end{itemize}
  \end{exampleblock}
  \pause
  \begin{itemize}
    \item[\faThumbsODown] Propositional model counting (\structure{$\#\SAT$}) is
          \alert{$\#\P$-complete}
    \item[\faThumbsOUp] But many of these counting problems have
          \alert{efficient solutions}
    \item And we can find them using \alert{first-order model counting}
          \begin{itemize}
            \item i.e., reasoning about sets, subsets, and arbitrary elements
                  without \alert{grounding} them
          \end{itemize}
  \end{itemize}
\end{frame}

\begin{frame}{More Formally: What Is the Input?}
  \begin{exampleblock}{Example Input Sentence}
    \[
      \forall x \in \alert<2>{\Gamma}\text{.
      }\forall y, z \in \alert<2>{\Delta}\text{.
      }P(x, y) \land P(x, z) \Rightarrow \alert<4>{y=z}
    \]
  \end{exampleblock}
  \begin{block}{\mbox{\alert<2>{Many-Sorted} \alert<3>{Function-Free}
        First-Order Logic with \alert<4>{Equality}}}
  \begin{itemize}
    \item<5-> Any number of variables and constants
    \item<5-> \structure{$\exists$} and \structure{$\forall$} quantifiers can be
          nested arbitrarily deeply
    \item<5-> All domains are finite
          \begin{itemize}
            \item Solutions are functions that take domain sizes as input
          \end{itemize}
    \item<5-> Of course, not all valid inputs have tractable solutions
  \end{itemize}
  \end{block}
  \onslide<6->{
    \begin{block}{First-Order Model Counting (FOMC)}
      \begin{itemize}
        \item Each predicate acts like a \alert{subset}
        \begin{itemize}
          \item of a domain or product of domains
        \end{itemize}
        \item Goal: count \alert{combinations of subsets} that satisfy the
              sentence
      \end{itemize}
    \end{block}
  }
\end{frame}

\begin{frame}{Exact Algorithms for FOMC}
  \begin{block}{Predecessors of This Work}
    \begin{itemize}
      \item
            \alert{\textsc{ForcLift}}~\textcolor{gray}{\parencite{DBLP:conf/ijcai/BroeckTMDR11}}
            \begin{itemize}
              \item knowledge compilation to \alert{FO d-DNNF}
            \end{itemize}
      \item \alert{\textsc{Crane}}~\textcolor{gray}{\parencite{DBLP:conf/kr/DilkasB23}}
            \begin{itemize}
              \item knowledge compilation to $\text{FO
                    d-DNNF} + \alert{\text{directed cycles}}$
              \item extends \alert{\textsc{ForcLift}} with support for:
                    \begin{itemize}
                      \item more input sentences and
                      \item recursive solutions
                    \end{itemize}
            \end{itemize}
    \end{itemize}
  \end{block}
  \pause
  \begin{block}{Other Approaches}
    \begin{itemize}
      \item \alert{L2C}~\textcolor{gray}{\parencite{DBLP:conf/kr/KazemiP16}}
            \begin{itemize}
              \item knowledge compilation to \alert{C++} code
            \end{itemize}
      \item
            \alert{Alchemy}~\textcolor{gray}{\parencite{DBLP:journals/cacm/GogateD16}}
            \begin{itemize}
              \item \alert{DPLL}-style search
            \end{itemize}
      \item
            \alert{FastWFOMC}~\textcolor{gray}{\parencite{DBLP:conf/uai/BremenK21}}
            \begin{itemize}
              \item based on \alert{cell enumeration}
              % \item with extensions for \alert{additional axioms}, ordered domains, etc.
            \end{itemize}
    \end{itemize}
  \end{block}
\end{frame}

\begin{frame}{Previous Work:
    \textsc{Crane}~\textcolor{gray}{\parencite{DBLP:conf/kr/DilkasB23}}}
  \begin{itemize}
    \item A knowledge compilation approach:
          \begin{itemize}
            \item \alert{Sentences} $\to$ labelled \alert{digraphs} $\to$
                  function-defining \alert{equations}
          \end{itemize}
    % \item Capable of constructing recursive solutions
    \item Two variants: \alert{greedy} search and \alert{breadth-first search}
          (BFS)
  \end{itemize}
  \pause
  \begin{exampleblock}{An Example Solution for Counting Bijections}
    \begin{align*}
      f(m, n) &= \sum_{l=0}^{n} \binom{n}{l}{(-1)}^{n-l}g(l, m),\\
      g(l, m) &= g(l-1, m) + mg(l-1, m-1)
    \end{align*}
  \end{exampleblock}
  \pause
  \begin{alertblock}{Issues We Are Going to Address}
    \begin{description}
      \item[Completeness:] recursive functions (like \structure{$g$}) have
            \alert{no base cases}
      \item[Usability:] how do I compute, e.g., \structure{$f(7, 7)$}?
    \end{description}
  \end{alertblock}
\end{frame}
% NOTE: the first but not the best: it can also find a linear-time solution
% NOTE: this slide formulates the problem tackled by this work

\begin{frame}{The Workflow of \textsc{Crane2} (1/2)}
  \begin{enumerate}
    \item Use \alert{\textsc{Crane}} to compile sentence \structure{$\phi$} into
          a set of equations \structure{$\mathcal{E}$} \pause
    \item Simplify them, e.g.,
          \[
            g(l, m) = \sum_{k=0}^{m}[0 \le k \le 1]\binom{m}{k}g(l-1, m-k)
          \]
          becomes
          \[
            g(l, m) = g(l-1, m) + mg(l-1, m-1)
          \] \pause
    \item[3. ($\Rightarrow$)] Identify a sufficient set of base cases
          \begin{itemize}
            \item e.g., \structure{$\{\, g(0, m), g(l, 0) \,\}$}
          \end{itemize}
  \end{enumerate}
\end{frame}
% NOTE: explain what the arrow means

@@

\begin{frame}[fragile]{From a Base Case to a Sentence}
  \begin{block}{From Previous
      Work~\textcolor{gray}{\parencite{DBLP:conf/kr/DilkasB23}}}
    \begin{itemize}
      \item \alert{\textsc{Crane}} associates each function \structure{$f$} with a
            sentence \structure{$\phi$} such that
            \structure{$\textsc{Crane}(\phi)$} produces the definition of
            \structure{$f$}
      \item There is a bijection between the parameters of the \alert{main
            function} \structure{$f$} and the \alert{domains} of the
            \alert{input sentence} \structure{$\phi$}
    \end{itemize}
  \end{block}
  \begin{example}
    \begin{itemize}
      \item Base case: \structure{$g(\stackrel{\mbox{\textcolor{black}{\tiny\onslide<2->{\ensuremath{\Gamma}}}}}{\stackrel{\mbox{\textcolor{black}{\tiny\onslide<2->{\ensuremath{\updownarrow}}}}}{0}}, \stackrel{\mbox{\textcolor{black}{\tiny\onslide<2->{\ensuremath{\Delta}}}}}{\stackrel{\mbox{\textcolor{black}{\tiny\onslide<2->{\ensuremath{\updownarrow}}}}}{m}})$}
      \item<3-> \alert{Part} of the sentence of \structure{$g$}:
            \begin{equation}\label{eq}
              \forall x \in \Gamma\text{. }\forall y \in \Delta\text{. }S(y) \lor \neg P(x, y)
            \end{equation}
      \item<4-> \structure{$g(0, \dots)$} means we need to simplify \eqref{eq}
            by assuming \structure{$|\Gamma| = 0$}
      \item<5-> Result: \structure{$\forall y \in \Delta\text{.
            }S(y) \lor \neg S(y)$}\hfill(\alert{Smoothing})
    \end{itemize}
  \end{example}
\end{frame}

\begin{frame}{An Outline of the Resulting C++ Program}
  \begin{algorithm}[H]
    initialise $\Cache_{\alert{g(0, m)}}$, $\Cache_{\alert{g(l, 0)}}$, $\Cache_{\alert{g}}$, and $\Cache_{\alert{f}}$\; \DontPrintSemicolon\pause
    \textbf{Function} $\alert{g_{0,m}(m)}$\textbf{:} \dots\;
    \textbf{Function} $\alert{g_{l,0}(l)}$\textbf{:} \dots\; \PrintSemicolon\pause
    \Fn{$\alert{g(l, m)}$}{
      \lIf{$(l, m) \in \Cache_{\alert{g}}$}{\Return{$\Cache_{\alert{g}}(l, m)$}}\label{line:cache}
      \lIf{$l = 0$}{\Return{$\alert{g_{0,m}(m)}$}}\label{line:basecaseone}
      \lIf{$m = 0$}{\Return{$\alert{g_{l,0}(l)}$}}\label{line:basecasetwo}
      $r \gets \alert{g(l-1, m)} + m\alert{g(l-1,m-1)}$\;
      $\Cache_{\alert{g}}(l, m) \gets r$\;
      \Return{$r$}\;
    } \DontPrintSemicolon
    \textbf{Function} $\alert{f(m, n)}$\textbf{:} \dots\; \PrintSemicolon
    \Fn{\alert{\Main}}{
      $(m, n) \gets \Input{}$\;
      \Return{$\alert{f(m, n)}$}\;
    }
  \end{algorithm}
\end{frame}

>>>>

\begin{frame}{Motivation}
  \begin{exampleblock}{Example Setting}
    \begin{itemize}
      \item Let \structure{$\Delta$} be a set of cardinality
            \structure{$n \in \mathbb{N}_0$}
      \item Suppose we want to count all \structure{$P \subseteq \Delta^{2}$}
            (as a function of \structure{$n$}) that are:
            \begin{itemize}
              \item functions,
              \item bijections,
              \item partial orders,
              \item symmetric,
              \item transitive,
              \item etc.
            \end{itemize}
    \end{itemize}
  \end{exampleblock}
  \pause
  \begin{itemize}
    \item[\faThumbsODown] Propositional model counting (\alert{$\#\SAT$}) is
          \alert{$\#\P$-complete}
    \item[\faThumbsOUp] But many of these counting problems have
          \alert{efficient solutions}
    \item And we can find them using \alert{first-order model counting}
          \begin{itemize}
            \item i.e., reasoning about sets, subsets, and arbitrary elements
                  without \alert{grounding} them
          \end{itemize}
  \end{itemize}
\end{frame}

\begin{frame}{More Formally: What Is the Input?}
  \begin{exampleblock}{Example Input Sentence}
    \[
      \forall x \in \alert<2>{\Gamma}\text{.
      }\forall y, z \in \alert<2>{\Delta}\text{.
      }P(x, y) \land P(x, z) \Rightarrow \alert<4>{y=z}
    \]
  \end{exampleblock}
  \begin{block}{\mbox{\alert<2>{Many-Sorted} \alert<3>{Function-Free}
        First-Order Logic with \alert<4>{Equality}}}
  \begin{itemize}
    \item<5-> Any number of variables and constants
    \item<5-> \structure{$\exists$} and \structure{$\forall$} quantifiers can be
          nested arbitrarily deeply
    \item<5-> All domains are finite
          \begin{itemize}
            \item Solutions are functions that take domain sizes as inputs
          \end{itemize}
    \item<5-> Of course, not all valid inputs have tractable solutions
  \end{itemize}
  \end{block}
  \onslide<6->{
    \begin{block}{First-Order Model Counting (FOMC)}
      \begin{itemize}
        \item Each predicate acts like a \alert{subset}
        \begin{itemize}
          \item of a domain or a Cartesian product of domains
        \end{itemize}
        \item Goal: count \alert{combinations of subsets} that satisfy the
              sentence
      \end{itemize}
    \end{block}
  }
\end{frame}

\begin{frame}{Exact Algorithms for FOMC}
  \begin{block}{Predecessors of This Work}
    \begin{itemize}
      \item
            \alert{\textsc{ForcLift}}~\textcolor{gray}{\parencite{DBLP:conf/ijcai/BroeckTMDR11}}
            \begin{itemize}
              \item knowledge compilation to \alert{FO d-DNNF}
            \end{itemize}
      \item \alert{\textsc{Crane}}~\textcolor{gray}{\parencite{DBLP:conf/kr/DilkasB23}}
            \begin{itemize}
              \item knowledge compilation to $\text{FO
                    d-DNNF} + \alert{\text{directed cycles}}$
              \item extends \alert{\textsc{ForcLift}} with support for:
                    \begin{itemize}
                      \item more input sentences and
                      \item recursive solutions
                    \end{itemize}
            \end{itemize}
    \end{itemize}
  \end{block}
  \pause
  \begin{block}{Some Other Approaches}
    \begin{itemize}
      \item \alert{L2C}~\textcolor{gray}{\parencite{DBLP:conf/kr/KazemiP16}}
            \begin{itemize}
              \item knowledge compilation to \alert{C++} code
            \end{itemize}
      \item
            \alert{Alchemy}~\textcolor{gray}{\parencite{DBLP:journals/cacm/GogateD16}}
            \begin{itemize}
              \item \alert{DPLL}-style search
            \end{itemize}
      \item
            \alert{FastWFOMC}~\textcolor{gray}{\parencite{DBLP:conf/uai/BremenK21}}
            \begin{itemize}
              \item based on \alert{cell enumeration}
              % \item with extensions for \alert{additional axioms}, ordered domains, etc.
            \end{itemize}
    \end{itemize}
  \end{block}
\end{frame}

\begin{frame}{Previous Work:
    \textsc{Crane}~\textcolor{gray}{\parencite{DBLP:conf/kr/DilkasB23}}}
  \begin{itemize}
    \item A knowledge compilation approach:
          \begin{itemize}
            \item \alert{Sentences} $\to$ labelled \alert{digraphs} $\to$
                  function-defining \alert{equations}
          \end{itemize}
    \item Two variants: \alert{greedy} search and \alert{breadth-first} search
          (BFS)
  \end{itemize}
  \pause
  \begin{exampleblock}{An Example Solution for Counting Bijections}
    \begin{align*}
      f(m, n) &= \sum_{l=0}^{n} \binom{n}{l}{(-1)}^{n-l}g(l, m),\\
      g(l, m) &= g(l-1, m) + mg(l-1, m-1)
    \end{align*}
  \end{exampleblock}
  \pause
  \begin{alertblock}{Issues We Are Going to Address}
    \begin{description}
      \item[Completeness:] recursive functions (like \structure{$g$}) have
            \alert{no base cases}
      \item[Usability:] how do I compute, e.g., \structure{$f(7, 7)$}?
    \end{description}
  \end{alertblock}
\end{frame}
% NOTE: the first but not the best: it can also find a linear-time solution
% NOTE: this slide formulates the problem tackled by this work

\begin{frame}{The Workflow of \textsc{Crane2} (1/2)}
  \begin{enumerate}
    \item Use \alert{\textsc{Crane}} to compile sentence \structure{$\phi$} into
          a set of equations \structure{$\mathcal{E}$} \pause
    \item Simplify them, e.g.,
          \[
            g(l, m) = \sum_{k=0}^{m}[0 \le k \le 1]\binom{m}{k}g(l-1, m-k)
          \]
          becomes
          \[
            g(l, m) = g(l-1, m) + mg(l-1, m-1)
          \] \pause
    \item[3. ($\Rightarrow$)] Identify a \alert{sufficient set of base cases} of
          all recursive functions
          \begin{itemize}
            \item e.g., \structure{$\{\, g(0, m), g(l, 0) \,\}$}
          \end{itemize}
  \end{enumerate}
\end{frame}
% NOTE: explain what the arrow means

@@

\begin{frame}[fragile]{From a Base Case to a Sentence}
  \begin{block}{From Previous
      Work~\textcolor{gray}{\parencite{DBLP:conf/kr/DilkasB23}}}
    \begin{itemize}
      \item \alert{\textsc{Crane}} associates each function \structure{$f$} with a
            sentence \structure{$\phi$} such that
            \structure{$\textsc{Crane}(\phi)$} produces the definition of
            \structure{$f$}
      \item And there is a bijection between the \alert{parameters} of
            \structure{$f$} and the \alert{domains} of \structure{$\phi$}
    \end{itemize}
  \end{block}
  \begin{example}
    \begin{itemize}
      \item Base case: \structure{$g(\stackrel{\mbox{\textcolor{black}{\tiny\onslide<2->{\ensuremath{\Gamma}}}}}{\stackrel{\mbox{\textcolor{black}{\tiny\onslide<2->{\ensuremath{\updownarrow}}}}}{0}}, \stackrel{\mbox{\textcolor{black}{\tiny\onslide<2->{\ensuremath{\Delta}}}}}{\stackrel{\mbox{\textcolor{black}{\tiny\onslide<2->{\ensuremath{\updownarrow}}}}}{m}})$}
      \item<3-> \alert{Part} of the sentence of \structure{$g$}:
            \begin{equation}\label{eq}
              \forall x \in \Gamma\text{. }\forall y \in \Delta\text{. }S(y) \lor \neg P(x, y)
            \end{equation}
      \item<4-> \structure{$g(0, \dots)$} means we need to simplify \eqref{eq}
            by assuming \structure{$|\Gamma| = 0$}
      \item<5-> Result: \structure{$\forall y \in \Delta\text{.
            }S(y) \lor \neg S(y)$}\hfill(\alert{Smoothing})
    \end{itemize}
  \end{example}
\end{frame}

\begin{frame}{The Structure of the Resulting C++ Program}
  \begin{algorithm}[H]
    initialise $\Cache_{\alert{g(0, m)}}$, $\Cache_{\alert{g(l, 0)}}$, $\Cache_{\alert{g}}$, and $\Cache_{\alert{f}}$\; \DontPrintSemicolon\pause
    \textbf{Function} $\alert{g_{0,m}(m)}$\textbf{:} \dots\;
    \textbf{Function} $\alert{g_{l,0}(l)}$\textbf{:} \dots\; \PrintSemicolon\pause
    \Fn{$\alert{g(l, m)}$}{
      \lIf{$(l, m) \in \Cache_{\alert{g}}$}{\Return{$\Cache_{\alert{g}}(l, m)$}}\label{line:cache}
      \lIf{$l = 0$}{\Return{$\alert{g_{0,m}(m)}$}}\label{line:basecaseone}
      \lIf{$m = 0$}{\Return{$\alert{g_{l,0}(l)}$}}\label{line:basecasetwo}
      $r \gets \alert{g(l-1, m)} + m\alert{g(l-1,m-1)}$\;
      $\Cache_{\alert{g}}(l, m) \gets r$\;
      \Return{$r$}\;
    } \DontPrintSemicolon
    \textbf{Function} $\alert{f(m, n)}$\textbf{:} \dots\; \PrintSemicolon
    \Fn{\alert{\Main}}{
      $(m, n) \gets \Input{}$\;
      \Return{$\alert{f(m, n)}$}\;
    }
  \end{algorithm}
\end{frame}
